---
title: "ggbeeswarm README examples"
subtitle: "p9beeswarm and ggbeeswarm side by side"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
execute:
  fig-width: 7
  fig-height: 5
---

This page recreates the [upstream ggbeeswarm README examples](https://github.com/eclarke/ggbeeswarm#examples).
The Python and upstream R plots are executed in adjacent Quarto cells and
displayed side by side. Click **Code** to expand either implementation.

## Rendering

Render from the repository root:

```sh
quarto render docs/ggbeeswarm-examples.qmd
```

Rendering requires R packages `ggplot2`, `ggbeeswarm`, and `reticulate`, plus
the project Python environment created by `uv sync --extra test`.

```{r}
#| include: false

library(ggplot2)
library(ggbeeswarm)
repo_root <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), ".."))
Sys.setenv(RETICULATE_PYTHON = file.path(repo_root, ".venv", "bin", "python"))
set.seed(12345)
alternative <- data.frame(x = "A", y = sample(1:100, 200, replace = TRUE))
priority <- data.frame(x = rep(1:3, c(20, 40, 80)))
priority$y <- rnorm(nrow(priority), priority$x)
set.seed(1995)
corral <- data.frame(y = rnorm(1000), id = sample(c("G1", "G2", "G3"), 1000, TRUE))
sub_mpg <- mpg[mpg$class %in% c("midsize", "pickup", "suv"), ]
```

```{python}
#| include: false

import io
import matplotlib.pyplot as plt
import numpy as np
import pandas as pd
import plotnine as p9
from plotnine.data import mpg
from sklearn.datasets import load_iris
from p9beeswarm import geom_beeswarm, geom_quasirandom

dataset = load_iris()
iris = pd.DataFrame(dataset.data, columns=["Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width"]).assign(Species=[dataset.target_names[i] for i in dataset.target])
sub_mpg = mpg[mpg["class"].isin(["midsize", "pickup", "suv"])]
rng = np.random.default_rng(12345)
alternative = pd.DataFrame({"x": "A", "y": rng.choice(np.arange(1, 101), 200)})
priority = pd.DataFrame({"x": np.repeat([1, 2, 3], [20, 40, 80])})
priority["y"] = rng.normal(priority["x"], 1)
corral_rng = np.random.default_rng(1995)
corral = pd.DataFrame({"y": corral_rng.normal(size=1000), "id": corral_rng.choice(["G1", "G2", "G3"], 1000)})

def show(plot):
    figure = plot.draw()
    figure.set_size_inches(7, 5)
    buffer = io.BytesIO()
    figure.savefig(buffer, format="png")
    buffer.seek(0)
    image = plt.imread(buffer)
    height, width = image.shape[:2]
    display_figure = plt.figure(figsize=(width / figure.dpi, height / figure.dpi), dpi=figure.dpi)
    display_figure.figimage(image, xo=0, yo=0)
    plt.show()
    plt.close("all")
```

## Jitter and quasirandom

:::: {.columns}
::: {.column width="50%"}
```{python}
#| code-summary: "Python: jitter and quasirandom"

show(p9.ggplot(iris, p9.aes("Species", "Sepal.Length")) + p9.geom_jitter())
show(p9.ggplot(iris, p9.aes("Species", "Sepal.Length")) + geom_quasirandom(random_state=12345))
```
:::
::: {.column width="50%"}
```{r}
#| code-summary: "R: jitter and quasirandom"

print(ggplot(iris, aes(Species, Sepal.Length)) + geom_jitter())
print(ggplot(iris, aes(Species, Sepal.Length)) + geom_quasirandom())
```
:::
::::

## geom_quasirandom

### Basic usage

:::: {.columns}
::: {.column width="50%"}
```{python}
#| code-summary: "Python: geom_quasirandom basic usage"

show(p9.ggplot(mpg, p9.aes("class", "hwy")) + geom_quasirandom(random_state=12345))
show(p9.ggplot(mpg, p9.aes("hwy", "class")) + geom_quasirandom(group_on_x=False, random_state=12345))
show(p9.ggplot(mpg, p9.aes("class", "hwy")) + geom_quasirandom(varwidth=True, random_state=12345))
show(p9.ggplot(sub_mpg, p9.aes("class", "displ", color="factor(cyl)")) + geom_quasirandom(dodge_width=1, random_state=12345))
```
:::
::: {.column width="50%"}
```{r}
#| code-summary: "R: geom_quasirandom basic usage"

print(ggplot(mpg, aes(class, hwy)) + geom_quasirandom())
print(ggplot(mpg, aes(hwy, class)) + geom_quasirandom(groupOnX = FALSE))
print(ggplot(mpg, aes(class, hwy)) + geom_quasirandom(varwidth = TRUE))
print(ggplot(sub_mpg, aes(class, displ, color = factor(cyl))) + geom_quasirandom(dodge.width = 1))
```
:::
::::

### Alternative methods

:::: {.columns}
::: {.column width="50%"}
```{python}
#| code-summary: "Python: geom_quasirandom alternative methods"

for method, title in [("tukey", "Tukey texture"), ("tukeyDense", "Tukey + density"), ("frowney", "Banded frowns"), ("smiley", "Banded smiles"), ("pseudorandom", "Jittered density")]:
    show(p9.ggplot(iris, p9.aes("Species", "Sepal.Length")) + geom_quasirandom(method=method, random_state=12345) + p9.ggtitle(title))
```
:::
::: {.column width="50%"}
```{r}
#| code-summary: "R: geom_quasirandom alternative methods"

for (method in c("tukey", "tukeyDense", "frowney", "smiley", "pseudorandom")) {
  print(ggplot(iris, aes(Species, Sepal.Length)) + geom_quasirandom(method = method) + ggtitle(method))
}
```
:::
::::

## geom_beeswarm

### Basic usage

:::: {.columns}
::: {.column width="50%"}
```{python}
#| code-summary: "Python: geom_beeswarm basic usage"

show(p9.ggplot(iris, p9.aes("Species", "Sepal.Length")) + geom_beeswarm())
show(p9.ggplot(iris, p9.aes("Species", "Sepal.Length")) + geom_beeswarm(side=1))
show(p9.ggplot(mpg, p9.aes("class", "hwy")) + geom_beeswarm(size=0.5))
show(p9.ggplot(mpg, p9.aes("hwy", "class")) + geom_beeswarm(size=0.5))
show(p9.ggplot(mpg, p9.aes("hwy", "class")) + geom_beeswarm(size=0.5) + p9.scale_y_discrete(expand=(0, 0, .5, 1)))
show(p9.ggplot(mpg, p9.aes("class", "hwy")) + geom_beeswarm(size=1.1))
show(p9.ggplot(sub_mpg, p9.aes("class", "displ", color="factor(cyl)")) + geom_beeswarm(dodge_width=.5))
```
:::
::: {.column width="50%"}
```{r}
#| code-summary: "R: geom_beeswarm basic usage"

print(ggplot(iris, aes(Species, Sepal.Length)) + geom_beeswarm())
print(ggplot(iris, aes(Species, Sepal.Length)) + geom_beeswarm(side = 1L))
print(ggplot(mpg, aes(class, hwy)) + geom_beeswarm(size = .5))
print(ggplot(mpg, aes(hwy, class)) + geom_beeswarm(size = .5))
print(ggplot(mpg, aes(hwy, class)) + geom_beeswarm(size = .5) + scale_y_discrete(expand = expansion(add = c(.5, 1))))
print(ggplot(mpg, aes(class, hwy)) + geom_beeswarm(size = 1.1))
print(ggplot(sub_mpg, aes(class, displ, color = factor(cyl))) + geom_beeswarm(dodge.width = .5))
```
:::
::::

### Alternative methods and priority

:::: {.columns}
::: {.column width="50%"}
```{python}
#| code-summary: "Python: geom_beeswarm methods and priority"

for method in ["swarm", "compactswarm", "hex", "square", "center"]:
    show(p9.ggplot(alternative, p9.aes("x", "y")) + geom_beeswarm(cex=2.5, method=method) + p9.ggtitle(f'method = "{method}"'))
for priority_name in ["ascending", "descending", "density", "random"]:
    show(p9.ggplot(priority, p9.aes("x", "y")) + geom_beeswarm(cex=2, priority=priority_name, random_state=12345) + p9.ggtitle(priority_name.title()) + p9.scale_x_continuous(expand=(0, 0, .5, .5)))
```
:::
::: {.column width="50%"}
```{r}
#| code-summary: "R: geom_beeswarm methods and priority"

for (method in c("swarm", "compactswarm", "hex", "square", "center")) {
  print(ggplot(alternative, aes(x, y)) + geom_beeswarm(cex = 2.5, method = method) + ggtitle(method))
}
for (priority_name in c("ascending", "descending", "density", "random")) {
  print(ggplot(priority, aes(x, y)) + geom_beeswarm(cex = 2, priority = priority_name) + ggtitle(priority_name) + scale_x_continuous(expand = expansion(add = c(.5, .5))))
}
```
:::
::::

### Corral runaway points

:::: {.columns}
::: {.column width="50%"}
```{python}
#| code-summary: "Python: geom_beeswarm corralling"

for corral_name in ["none", "gutter", "wrap", "random", "omit"]:
    show(p9.ggplot(corral, p9.aes("id", "y", color="id")) + geom_beeswarm(cex=2.5, corral=corral_name, corral_width=.9, random_state=1995) + p9.ggtitle(f'corral = "{corral_name}"'))
```
:::
::: {.column width="50%"}
```{r}
#| code-summary: "R: geom_beeswarm corralling"

for (corral_name in c("none", "gutter", "wrap", "random", "omit")) {
  print(ggplot(corral, aes(id, y, colour = id)) + geom_beeswarm(cex = 2.5, corral = corral_name, corral.width = .9) + ggtitle(corral_name))
}
```
:::
::::
