---
title: "ggbeeswarm README examples"
subtitle: "p9beeswarm and ggbeeswarm side by side"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
execute:
  fig-width: 7
  fig-height: 5
---

This page recreates the [upstream ggbeeswarm README examples](https://github.com/eclarke/ggbeeswarm#examples).
The Python and upstream R plots are executed in adjacent Quarto cells and
displayed side by side. Click **Code** to expand either implementation.

## Rendering

Render from the repository root:

```sh
quarto render docs/ggbeeswarm-examples.qmd
```

Rendering requires R packages `ggplot2`, `ggbeeswarm`, and `reticulate`, plus
the project Python environment created by `uv sync --extra test`.

```{r}
#| include: false

library(ggplot2)
library(ggbeeswarm)
repo_root <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), ".."))
Sys.setenv(RETICULATE_PYTHON = file.path(repo_root, ".venv", "bin", "python"))
set.seed(12345)
alternative <- data.frame(x = "A", y = sample(1:100, 200, replace = TRUE))
priority <- data.frame(x = rep(1:3, c(20, 40, 80)))
priority$y <- rnorm(nrow(priority), priority$x)
set.seed(1995)
corral <- data.frame(y = rnorm(1000), id = sample(c("G1", "G2", "G3"), 1000, TRUE))
sub_mpg <- mpg[mpg$class %in% c("midsize", "pickup", "suv"), ]
```

```{python}
#| include: false

import numpy as np
import pandas as pd
import plotnine as p9
from plotnine.data import mpg
from sklearn.datasets import load_iris
from p9beeswarm import geom_beeswarm, geom_quasirandom

dataset = load_iris()
iris = pd.DataFrame(dataset.data, columns=["Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width"]).assign(Species=[dataset.target_names[i] for i in dataset.target])
sub_mpg = mpg[mpg["class"].isin(["midsize", "pickup", "suv"])]
rng = np.random.default_rng(12345)
alternative = pd.DataFrame({"x": "A", "y": rng.choice(np.arange(1, 101), 200)})
priority = pd.DataFrame({"x": np.repeat([1, 2, 3], [20, 40, 80])})
priority["y"] = rng.normal(priority["x"], 1)
corral_rng = np.random.default_rng(1995)
corral = pd.DataFrame({"y": corral_rng.normal(size=1000), "id": corral_rng.choice(["G1", "G2", "G3"], 1000)})
```

## Jitter and quasirandom

::: {layout-ncol=2}
```{python Python: jitter}
#| fig-cap: "Python: jitter"

(p9.ggplot(iris, p9.aes("Species", "Sepal.Length")) + p9.geom_jitter()).show()
```
```{r R: jitter}
#| fig-cap: "R: jitter"

print(ggplot(iris, aes(Species, Sepal.Length)) + geom_jitter())
```
```{python Python: quasirandom}
#| fig-cap: "Python: quasirandom"

(p9.ggplot(iris, p9.aes("Species", "Sepal.Length")) + geom_quasirandom(random_state=12345)).show()
```
```{r R: quasirandom}
#| fig-cap: "R: quasirandom"

print(ggplot(iris, aes(Species, Sepal.Length)) + geom_quasirandom())
```
:::

## geom_quasirandom

### Basic usage

::: {layout-ncol=2}
```{python Python: basic geom_quasirandom}
#| fig-cap: "Python: basic geom_quasirandom"

(p9.ggplot(mpg, p9.aes("class", "hwy")) + geom_quasirandom(random_state=12345)).show()
```
```{r R: basic geom_quasirandom}
#| fig-cap: "R: basic geom_quasirandom"

print(ggplot(mpg, aes(class, hwy)) + geom_quasirandom())
```
```{python Python: horizontal geom_quasirandom}
#| fig-cap: "Python: horizontal geom_quasirandom"

(p9.ggplot(mpg, p9.aes("hwy", "class")) + geom_quasirandom(group_on_x=False, random_state=12345)).show()
```
```{r R: horizontal geom_quasirandom}
#| fig-cap: "R: horizontal geom_quasirandom"

print(ggplot(mpg, aes(hwy, class)) + geom_quasirandom(groupOnX = FALSE))
```
```{python Python: varwidth geom_quasirandom}
#| fig-cap: "Python: varwidth geom_quasirandom"

(p9.ggplot(mpg, p9.aes("class", "hwy")) + geom_quasirandom(varwidth=True, random_state=12345)).show()
```
```{r R: varwidth geom_quasirandom}
#| fig-cap: "R: varwidth geom_quasirandom"

print(ggplot(mpg, aes(class, hwy)) + geom_quasirandom(varwidth = TRUE))
```
```{python Python: dodged geom_quasirandom}
#| fig-cap: "Python: dodged geom_quasirandom"

(p9.ggplot(sub_mpg, p9.aes("class", "displ", color="factor(cyl)")) + geom_quasirandom(dodge_width=1, random_state=12345)).show()
```
```{r R: dodged geom_quasirandom}
#| fig-cap: "R: dodged geom_quasirandom"

print(ggplot(sub_mpg, aes(class, displ, color = factor(cyl))) + geom_quasirandom(dodge.width = 1))
```
:::

### Alternative methods

::: {layout-ncol=2}
```{python Python: Tukey texture}
#| fig-cap: "Python: Tukey texture"

(p9.ggplot(iris, p9.aes("Species", "Sepal.Length")) + geom_quasirandom(method="tukey", random_state=12345) + p9.ggtitle("Tukey texture")).show()
```
```{r R: Tukey texture}
#| fig-cap: "R: Tukey texture"

print(ggplot(iris, aes(Species, Sepal.Length)) + geom_quasirandom(method = "tukey") + ggtitle("tukey"))
```
```{python Python: Tukey density}
#| fig-cap: "Python: Tukey density"

(p9.ggplot(iris, p9.aes("Species", "Sepal.Length")) + geom_quasirandom(method="tukeyDense", random_state=12345) + p9.ggtitle("Tukey + density")).show()
```
```{r R: Tukey density}
#| fig-cap: "R: Tukey density"

print(ggplot(iris, aes(Species, Sepal.Length)) + geom_quasirandom(method = "tukeyDense") + ggtitle("tukeyDense"))
```
```{python Python: frowney quasirandom}
#| fig-cap: "Python: frowney quasirandom"

(p9.ggplot(iris, p9.aes("Species", "Sepal.Length")) + geom_quasirandom(method="frowney", random_state=12345) + p9.ggtitle("Banded frowns")).show()
```
```{r R: frowney quasirandom}
#| fig-cap: "R: frowney quasirandom"

print(ggplot(iris, aes(Species, Sepal.Length)) + geom_quasirandom(method = "frowney") + ggtitle("frowney"))
```
```{python Python: smiley quasirandom}
#| fig-cap: "Python: smiley quasirandom"

(p9.ggplot(iris, p9.aes("Species", "Sepal.Length")) + geom_quasirandom(method="smiley", random_state=12345) + p9.ggtitle("Banded smiles")).show()
```
```{r R: smiley quasirandom}
#| fig-cap: "R: smiley quasirandom"

print(ggplot(iris, aes(Species, Sepal.Length)) + geom_quasirandom(method = "smiley") + ggtitle("smiley"))
```
```{python Python: pseudorandom quasirandom}
#| fig-cap: "Python: pseudorandom quasirandom"

(p9.ggplot(iris, p9.aes("Species", "Sepal.Length")) + geom_quasirandom(method="pseudorandom", random_state=12345) + p9.ggtitle("Jittered density")).show()
```
```{r R: pseudorandom quasirandom}
#| fig-cap: "R: pseudorandom quasirandom"

print(ggplot(iris, aes(Species, Sepal.Length)) + geom_quasirandom(method = "pseudorandom") + ggtitle("pseudorandom"))
```
:::

## geom_beeswarm

### Basic usage

::: {layout-ncol=2}
```{python Python: basic geom_beeswarm}
#| fig-cap: "Python: basic geom_beeswarm"

(p9.ggplot(iris, p9.aes("Species", "Sepal.Length")) + geom_beeswarm()).show()
```
```{r R: basic geom_beeswarm}
#| fig-cap: "R: basic geom_beeswarm"

print(ggplot(iris, aes(Species, Sepal.Length)) + geom_beeswarm())
```
```{python Python: one-sided geom_beeswarm}
#| fig-cap: "Python: one-sided geom_beeswarm"

(p9.ggplot(iris, p9.aes("Species", "Sepal.Length")) + geom_beeswarm(side=1)).show()
```
```{r R: one-sided geom_beeswarm}
#| fig-cap: "R: one-sided geom_beeswarm"

print(ggplot(iris, aes(Species, Sepal.Length)) + geom_beeswarm(side = 1L))
```
```{python Python: compact vertical geom_beeswarm}
#| fig-cap: "Python: compact vertical geom_beeswarm"

(p9.ggplot(mpg, p9.aes("class", "hwy")) + geom_beeswarm(size=0.5)).show()
```
```{r R: compact vertical geom_beeswarm}
#| fig-cap: "R: compact vertical geom_beeswarm"

print(ggplot(mpg, aes(class, hwy)) + geom_beeswarm(size = .5))
```
```{python Python: compact horizontal geom_beeswarm}
#| fig-cap: "Python: compact horizontal geom_beeswarm"

(p9.ggplot(mpg, p9.aes("hwy", "class")) + geom_beeswarm(size=0.5)).show()
```
```{r R: compact horizontal geom_beeswarm}
#| fig-cap: "R: compact horizontal geom_beeswarm"

print(ggplot(mpg, aes(hwy, class)) + geom_beeswarm(size = .5))
```
```{python Python: expanded horizontal geom_beeswarm}
#| fig-cap: "Python: expanded horizontal geom_beeswarm"

(p9.ggplot(mpg, p9.aes("hwy", "class")) + geom_beeswarm(size=0.5) + p9.scale_y_discrete(expand=(0, 0, .5, 1))).show()
```
```{r R: expanded horizontal geom_beeswarm}
#| fig-cap: "R: expanded horizontal geom_beeswarm"

print(ggplot(mpg, aes(hwy, class)) + geom_beeswarm(size = .5) + scale_y_discrete(expand = expansion(add = c(.5, 1))))
```
```{python Python: large-point geom_beeswarm}
#| fig-cap: "Python: large-point geom_beeswarm"

(p9.ggplot(mpg, p9.aes("class", "hwy")) + geom_beeswarm(size=1.1)).show()
```
```{r R: large-point geom_beeswarm}
#| fig-cap: "R: large-point geom_beeswarm"

print(ggplot(mpg, aes(class, hwy)) + geom_beeswarm(size = 1.1))
```
```{python Python: dodged geom_beeswarm}
#| fig-cap: "Python: dodged geom_beeswarm"

(p9.ggplot(sub_mpg, p9.aes("class", "displ", color="factor(cyl)")) + geom_beeswarm(dodge_width=.5)).show()
```
```{r R: dodged geom_beeswarm}
#| fig-cap: "R: dodged geom_beeswarm"

print(ggplot(sub_mpg, aes(class, displ, color = factor(cyl))) + geom_beeswarm(dodge.width = .5))
```
:::

### Alternative methods and priority

::: {layout-ncol=2}
```{python Python: swarm method}
#| fig-cap: "Python: swarm method"

(p9.ggplot(alternative, p9.aes("x", "y")) + geom_beeswarm(cex=2.5, method="swarm") + p9.ggtitle('method = "swarm"')).show()
```
```{r R: swarm method}
#| fig-cap: "R: swarm method"

print(ggplot(alternative, aes(x, y)) + geom_beeswarm(cex = 2.5, method = "swarm") + ggtitle("swarm"))
```
```{python Python: compactswarm method}
#| fig-cap: "Python: compactswarm method"

(p9.ggplot(alternative, p9.aes("x", "y")) + geom_beeswarm(cex=2.5, method="compactswarm") + p9.ggtitle('method = "compactswarm"')).show()
```
```{r R: compactswarm method}
#| fig-cap: "R: compactswarm method"

print(ggplot(alternative, aes(x, y)) + geom_beeswarm(cex = 2.5, method = "compactswarm") + ggtitle("compactswarm"))
```
```{python Python: hex method}
#| fig-cap: "Python: hex method"

(p9.ggplot(alternative, p9.aes("x", "y")) + geom_beeswarm(cex=2.5, method="hex") + p9.ggtitle('method = "hex"')).show()
```
```{r R: hex method}
#| fig-cap: "R: hex method"

print(ggplot(alternative, aes(x, y)) + geom_beeswarm(cex = 2.5, method = "hex") + ggtitle("hex"))
```
```{python Python: square method}
#| fig-cap: "Python: square method"

(p9.ggplot(alternative, p9.aes("x", "y")) + geom_beeswarm(cex=2.5, method="square") + p9.ggtitle('method = "square"')).show()
```
```{r R: square method}
#| fig-cap: "R: square method"

print(ggplot(alternative, aes(x, y)) + geom_beeswarm(cex = 2.5, method = "square") + ggtitle("square"))
```
```{python Python: center method}
#| fig-cap: "Python: center method"

(p9.ggplot(alternative, p9.aes("x", "y")) + geom_beeswarm(cex=2.5, method="center") + p9.ggtitle('method = "center"')).show()
```
```{r R: center method}
#| fig-cap: "R: center method"

print(ggplot(alternative, aes(x, y)) + geom_beeswarm(cex = 2.5, method = "center") + ggtitle("center"))
```
```{python Python: ascending priority}
#| fig-cap: "Python: ascending priority"

(p9.ggplot(priority, p9.aes("x", "y")) + geom_beeswarm(cex=2, priority="ascending", random_state=12345) + p9.ggtitle("Ascending") + p9.scale_x_continuous(expand=(0, 0, .5, .5))).show()
```
```{r R: ascending priority}
#| fig-cap: "R: ascending priority"

print(ggplot(priority, aes(x, y)) + geom_beeswarm(cex = 2, priority = "ascending") + ggtitle("ascending") + scale_x_continuous(expand = expansion(add = c(.5, .5))))
```
```{python Python: descending priority}
#| fig-cap: "Python: descending priority"

(p9.ggplot(priority, p9.aes("x", "y")) + geom_beeswarm(cex=2, priority="descending", random_state=12345) + p9.ggtitle("Descending") + p9.scale_x_continuous(expand=(0, 0, .5, .5))).show()
```
```{r R: descending priority}
#| fig-cap: "R: descending priority"

print(ggplot(priority, aes(x, y)) + geom_beeswarm(cex = 2, priority = "descending") + ggtitle("descending") + scale_x_continuous(expand = expansion(add = c(.5, .5))))
```
```{python Python: density priority}
#| fig-cap: "Python: density priority"

(p9.ggplot(priority, p9.aes("x", "y")) + geom_beeswarm(cex=2, priority="density", random_state=12345) + p9.ggtitle("Density") + p9.scale_x_continuous(expand=(0, 0, .5, .5))).show()
```
```{r R: density priority}
#| fig-cap: "R: density priority"

print(ggplot(priority, aes(x, y)) + geom_beeswarm(cex = 2, priority = "density") + ggtitle("density") + scale_x_continuous(expand = expansion(add = c(.5, .5))))
```
```{python Python: random priority}
#| fig-cap: "Python: random priority"

(p9.ggplot(priority, p9.aes("x", "y")) + geom_beeswarm(cex=2, priority="random", random_state=12345) + p9.ggtitle("Random") + p9.scale_x_continuous(expand=(0, 0, .5, .5))).show()
```
```{r R: random priority}
#| fig-cap: "R: random priority"

print(ggplot(priority, aes(x, y)) + geom_beeswarm(cex = 2, priority = "random") + ggtitle("random") + scale_x_continuous(expand = expansion(add = c(.5, .5))))
```
:::

### Corral runaway points

::: {layout-ncol=2}
```{python Python: no corralling}
#| fig-cap: "Python: no corralling"

(p9.ggplot(corral, p9.aes("id", "y", color="id")) + geom_beeswarm(cex=2.5, corral="none", corral_width=.9, random_state=1995) + p9.ggtitle('corral = "none"')).show()
```
```{r R: no corralling}
#| fig-cap: "R: no corralling"

print(ggplot(corral, aes(id, y, colour = id)) + geom_beeswarm(cex = 2.5, corral = "none", corral.width = .9) + ggtitle("none"))
```
```{python Python: gutter corralling}
#| fig-cap: "Python: gutter corralling"

(p9.ggplot(corral, p9.aes("id", "y", color="id")) + geom_beeswarm(cex=2.5, corral="gutter", corral_width=.9, random_state=1995) + p9.ggtitle('corral = "gutter"')).show()
```
```{r R: gutter corralling}
#| fig-cap: "R: gutter corralling"

print(ggplot(corral, aes(id, y, colour = id)) + geom_beeswarm(cex = 2.5, corral = "gutter", corral.width = .9) + ggtitle("gutter"))
```
```{python Python: wrapped corralling}
#| fig-cap: "Python: wrapped corralling"

(p9.ggplot(corral, p9.aes("id", "y", color="id")) + geom_beeswarm(cex=2.5, corral="wrap", corral_width=.9, random_state=1995) + p9.ggtitle('corral = "wrap"')).show()
```
```{r R: wrapped corralling}
#| fig-cap: "R: wrapped corralling"

print(ggplot(corral, aes(id, y, colour = id)) + geom_beeswarm(cex = 2.5, corral = "wrap", corral.width = .9) + ggtitle("wrap"))
```
```{python Python: random corralling}
#| fig-cap: "Python: random corralling"

(p9.ggplot(corral, p9.aes("id", "y", color="id")) + geom_beeswarm(cex=2.5, corral="random", corral_width=.9, random_state=1995) + p9.ggtitle('corral = "random"')).show()
```
```{r R: random corralling}
#| fig-cap: "R: random corralling"

print(ggplot(corral, aes(id, y, colour = id)) + geom_beeswarm(cex = 2.5, corral = "random", corral.width = .9) + ggtitle("random"))
```
```{python Python: omitted corralling}
#| fig-cap: "Python: omitted corralling"

(p9.ggplot(corral, p9.aes("id", "y", color="id")) + geom_beeswarm(cex=2.5, corral="omit", corral_width=.9, random_state=1995) + p9.ggtitle('corral = "omit"')).show()
```
```{r R: omitted corralling}
#| fig-cap: "R: omitted corralling"

print(ggplot(corral, aes(id, y, colour = id)) + geom_beeswarm(cex = 2.5, corral = "omit", corral.width = .9) + ggtitle("omit"))
```
:::
